---
title: "EDA_data_process_Candice"
author: "Candice Yu"
date: "2024-03-23"
output:
    pdf_document:
       latex_engine: xelatex
       toc: true
       toc_depth: 2
---
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r echo = T, message = FALSE, results = 'hide', warning = FALSE}
library(caret)
library(tidyverse)
library(gridExtra)
```

# Load Data 
```{r}
load("./Data/recovery.RData")
dat <- dat %>%
  mutate(gender = as_factor(gender),
         diabetes = as_factor(diabetes),
         hypertension = as_factor(hypertension),
         vaccine = as_factor(vaccine),
         severity = as_factor(severity)) %>%
  select(-id)
```


# EDA
## Overview of the Data 
```{r}
# brief summary of the data 
skimr::skim(dat)
```

## EDA for Continuous Variables 

### Correlation plot for continuous variables
```{r}
# correlation plot for continuous variables
continuous_vars <- dat %>% 
  select(height, weight, bmi, SBP, LDL, recovery_time)
correlations <- cor(continuous_vars) 
corrplot::corrplot(correlations, method = "circle")
```

The correlation plot shows some relationships between variables, but none of them appear to be strongly correlated with `recovery_time`. This may suggest that linear relationships are not strong, and hence a non-linear model could be more appropriate.


### Scatter plots to explore potential relationships
```{r}
# between time to recovery and bmi
ggplot(dat, aes(x = bmi, y = recovery_time, color = gender)) + 
  geom_point() + geom_smooth(method = "lm") +
  labs(title = "Time to Recovery vs. BMI", 
       x = "BMI", 
       y = "Time to Recovery (Days)")

# between time to recovery and height
ggplot(dat, aes(x = height, y = recovery_time, color = gender)) + 
  geom_point() + geom_smooth(method = "lm") +
  labs(title = "Time to Recovery vs. Height", 
       x = "Height", 
       y = "Time to Recovery (Days)")
```
## EDA for Discrete Variables 
### Heatmap for severity and vaccination status
```{r}
# Heatmap for systolic blood pressure across severity and vaccination status
dat %>% 
  group_by(severity, vaccine) %>%
  summarise(avg_recovery_time = mean(recovery_time)) %>%
  ggplot(aes(x = factor(severity), y = factor(vaccine), fill = avg_recovery_time)) +
  geom_tile() +
  labs(x = "Severity of COVID-19", 
       y = "Vaccination Status", 
       fill = "Average Recovery Time")
```
### Faceted grid plot for categorical variables
```{r, fig.width = 8, fig.height = 10}

# faceted grid plot for categorical variables
categorical_vars <- c("smoking", "diabetes", "hypertension", "vaccine", "severity")
faceted_plots <- lapply(categorical_vars, function(var) {
  ggplot(dat, aes_string(x = var, y = "recovery_time")) + 
    geom_boxplot(aes(fill = factor(..x..))) + 
    facet_wrap(~study) + 
    labs(title = paste("Recovery Time by", var), y = "Time to Recovery") +
    theme_minimal()
})

# combine the plots into one grid
grid.arrange(scatter_bmi, scatter_sbp, scatter_ldl, grobs = faceted_plots, ncol = 2)
```

The boxplots indicate a significant difference in recovery times between study groups A and B across several categorical factors, which suggests that `study` is an important variable to include in the model.


# Preprocess of the Data 
```{r}
data <- dat

# normalize/standardize numerical variables 
num_vars <- names(data)[sapply(data, is.numeric)][-7]
preprocess_params <- preProcess(data[, num_vars], method = c("center", "scale"))
data[num_vars] <- predict(preprocess_params, data[, num_vars])

# log transform 'recovery_time' since it's highly skewed
data$recovery_time <- log(data$recovery_time) 
```


## Split data & Define the control method
```{r}
# split data into training and test sets
set.seed(123) 
indexes <- createDataPartition(data$recovery_time, p = 0.8, list = FALSE)
train_data <- data[indexes, ]
test_data <- data[-indexes, ]

# matrix of predictors
x <- train_data %>% select(-recovery_time)
y <- train_data$recovery_time

# define the control method for training
ctrl1 <- trainControl(method = "cv", number = 10) # 10-fold cross-validation

```

## Export the the training/test set & control method
```{r}
# save the training and test sets to CSV files
write.csv(train_data, "./Data/train_data.csv", row.names = FALSE)
write.csv(test_data, "./Data/test_data.csv", row.names = FALSE)

# save the control method using saveRDS
saveRDS(ctrl1, "./Data/train_control.rds")

```

## Load the training/test set & control method
```{r}
# Load the training and test sets
train_data <- read.csv("./Data/train_data.csv")
test_data <- read.csv("./Data/test_data.csv")

# Load the control method
ctrl1 <- readRDS("./Data/train_control.rds")

# change variables to be factors again
train_data <- train_data %>%
  mutate(gender = as_factor(gender),
         diabetes = as_factor(diabetes),
         hypertension = as_factor(hypertension),
         vaccine = as_factor(vaccine),
         severity = as_factor(severity))

test_data <- test_data %>%
  mutate(gender = as_factor(gender),
         diabetes = as_factor(diabetes),
         hypertension = as_factor(hypertension),
         vaccine = as_factor(vaccine),
         severity = as_factor(severity))
```


