---
title: "GAM"
date: "2024-03-21"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: 3
  header-includes:
    -\usepackage{fancyhdr}
    -\usepackage{lipsum}
    -\pagestyle{fancy}
    -\fancyhead[R]{\thepage}
    -\fancypagestyle{plain}{\pagestyle{fancy}}
---
```{r setup}
knitr::opts_chunk$set(
  collapse = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.dim = c(10, 5),
  fig.format = "png")
```

# Load Data and Package

```{r}
library(tidyverse)
library(caret)
## Load the the training/test set & control method

# Load the training and test sets
train_data <- read.csv("./Data/train_data.csv")
test_data <- read.csv("./Data/test_data.csv")

# Load the control method
ctrl1 <- readRDS("./Data/train_control.rds")

# change variables to be factors again
train_data <- train_data %>%
  mutate(gender = as_factor(gender),
         diabetes = as_factor(diabetes),
         hypertension = as_factor(hypertension),
         vaccine = as_factor(vaccine),
         severity = as_factor(severity))

test_data <- test_data %>%
  mutate(gender = as_factor(gender),
         diabetes = as_factor(diabetes),
         hypertension = as_factor(hypertension),
         vaccine = as_factor(vaccine),
         severity = as_factor(severity))
```

# GAM


- **Family and Link Function**: The model assumes a Gaussian family for the error distribution and uses an identity link function. This is typical for a regression problem where the outcome is continuous.

**Formula**: The outcome is modeled as a linear combination of several covariates, including gender, hypertension, diabetes, and others. Additionally, smooth terms for age, SBP, LDL, BMI, height, and weight are included to capture non-linear effects.

 **GCV Score**: The Generalized Cross-Validation (GCV) score is 0.1915381, which assesses the model's predictive performance, with lower values indicating better fit.

**Parametric Coefficients**: The table lists the estimated coefficients for each predictor, with their standard errors, t-values, and p-values. For instance, the coefficient for 'gender1' is -0.081585, and it is statistically significant at the 0.001 level (indicated by ***), suggesting a strong relationship with the outcome.

**Smooth Terms Significance**: This section presents the significance of smooth terms, with 's(LDL)' and 's(BMI)' being highly significant (p < 0.001), indicating important non-linear relationships with the outcome.

**Model Performance Metrics**:
  - The adjusted R-squared value is 0.195, indicating that approximately 19.5% of the variance in the outcome is explained by the model.
  - Deviance explained is 20.3%, which is another way of measuring model fit, similar to R-squared.
  - The scale estimate is about 0.18951.
  - The size of the dataset used for the model is 2,402 observations.

In summary, this GAM model includes both linear and non-linear relationships between predictors and the outcome. Some predictors show significant effects on the outcome, with non-linear relationships for variables like BMI and LDL being notably significant. Overall, the model explains a modest proportion of the variance in the outcome, suggesting there may be other factors not included in the model that also influence the outcome.


```{r}
set.seed(1)

x_train = train_data[1:14]
y_train = train_data$recovery_time

x_test = test_data[1:14]
y_test = test_data$recovery_time

model.gam <- train(x = x_train,
                   y = y_train,
                   method = "gam",
                   #metric = "RMSE", by default
                   trControl = trainControl(method = "cv", number = 10))



summary(model.gam$finalModel)
#plot(model.gam$finalModel)

# Calculate training RMSE of optimal model
gam_train_RMSE = sqrt(mean((y_train - predict(model.gam))^2))
gam_train_RMSE

# Calculate test RMSE of optimal model
test_predictions = predict(model.gam, x_test)

gam_test_RMSE = sqrt(mean((y_test - test_predictions)^2))
gam_test_RMSE

# Total Sum of Squares
TSS <- sum((train_data$recovery_time - mean(train_data$recovery_time))^2)

# Residual Sum of Squares
RSS <- sum((train_data$recovery_time - predict(model.gam))^2)

# R²
R2 <- 1 - (RSS/TSS)
R2
```
